---
title: "Berkeley Formal Demography Workshop 2019, Introductory Lab"
author: Joshua R. Goldstein
date: "17 May 2018"
output: html_notebook
---


# Overview

This is a minimal example for everyone to try to get started using R in the Rstudio nvironment inside an R notebook. 

Our goals in this first lab are:

1. To get everyone started programming and doing labs with R and RStudio.

2. To expose you to a tiny bit of fertility.

> This document is written in the "R markdown" format and should be
> read and edited within RStudio. The notebook interface allows us to
> execute and display R-code within a single window. You can edit this
> notebook directly and save it. We recommend you save it with a
> different name (e.g., "BFDW_2018_intro_lab_myname.Rmd") to avoid overwriting
> your edits.


Some of you may already know R or another programming language and have already used RStudio. If that's the case, just leap in and do this lab right away. Even if you've never used R or RStudio, it might still be a good idea to go through the lab first and then consult some of the following resources:

* https://cran.r-project.org/doc/contrib/Torfs+Brauer-Short-R-Intro.pdf

This is a quite thorough introduction, for those who like to read.

* Intro to R - YouTube
www.youtube.com/playlist?list=PLOU2XLYxmsIK9qQfztXeybpHvru-TrqAP

For those who like to sit back and watch ... A very nice set of short ( a few minutes each ) videos created by Google for their programmers. They don't use RStudio, so video 1 might be a bit foreign, but after that they are all great for you.

* Try R - Code School
http://tryr.codeschool.com/

For those who like to play ... try this interactive way to start out with R.


For RStudio and using R Notebooks (like this file), you probably will do best just walking your way through the notebook itself. There is a fair amount about this on the internet but it is often aimed at compiling documents or webpages from the notebooks, whereas we are just going to use the notebooks as a stand-alone. Our notebooks also have a cool feature of auto-graded quiz questions.

# Writing and executing R commands

(Note: In addition to this introduction, you may want to consult "The Rstudio175 cloud server for Demog/Econ C175" by Carl Mason available at the course website [http://courses.demog.berkeley.edu/formaldemography/], which provides additional information for a very similar environment we developed for Berkeley undergrads last semester.]

In the RStudio Notebook, we can write and execute R commands in the
same document that we are writing text in by creating a "code chunk."
To open a code chunk, we type ""```{r}" to start. To close the code
chunk we type "```"" to end. Here are some examples:

To print a string of text:
```{r}
print("Hello, world")
```
You can hit the green "play" icon to execute the chunk of code. Try modifying the chunk so it prints: "Hello, Berkeley". (Note the [1] just means that it is the first element of the displayed object. You can ignore this.)

To add 2 + 2:
```{r}
2+2
```

To assign the value 4 to a variable named "x":
```{r}
x <- 4
print(x)
```

The output of a chunk is only shown when you tell it to be. For example,
```{r}
x <- 3
print(x)
y <- 4
## note: we don't tell R to show us the value of "y"
z <- 5
print(z)
```


Let's calculate a fertility rate. Say there were 10 births to 45 women aged 20-25 in 1987.

```{r}
nBx = 10
nKx.female = 45
nFx = nBx/nKx.female
print(nFx)
```

R is a vectorized language. Here as an example using vectors, taken from Keyfitz and Flieger (1968, page 312) for 1851 France.

```{r}

## some data typed in by hand by Josh
million = 10^6 # to save us typing and to make code more readable
x <- c(15, 20, 25, 30, 35, 40) # ages at start of each interval, equivalent to 'seq(15, 40, 5)'
nKx.female.vec = c(1.55, 1.52, 1.44, 1.35, 1.27, 1.17) * million # suffix .vec just for us humans 
nBx.vec = c(39, 212, 283, 235, 147, 54) * 1000

## now calculate age-specific fertility rates
nFx.vec <- nBx.vec/nKx.female.vec   # can add, subtract, multiply and divide vectors 
## (Note this is element by element and not matrix multiplication, which has a different syntax)

## Plot the result
plot(x = x, y = nFx.vec, 
     type = "o",   # plotting type "o" means points _o_ver lines (try changing to "p", or "l", or "b" and reclicking the 'play' button for this chunk)
     ylim = c(0, .25))
title("Period Age-specific Fertility in France, 1851")
```


Let's do one more example, calculating the TFR in France in 1851. The objects carry over from chunk to chunk. So we can just pick up where we left off ...

```{r}
n = rep(x = 5, times = length(x)) ## a vector of 5s of the right length
TFR = sum(n * nFx.vec) ## sum of the vector that results from the product of n * nFx.vec
print(TFR)
```


Some exercises:


1. Create a new chunk and use R to calculate the NRR using the approximation

$$
NRR = TFR \cdot \ell_\mu \cdot p_{female} 
$$
Here's some skeletal code
```{r}
NRR = TFR * addsomestuffhere
print(NRR)
```

with the survival to the mean age of childbearing $\ell_\mu = 0.58$, the mean age of childbearing $\mu = 29.7$, and the proportion female at birth $p_{female} = 0.4886. (Note if you play with your mouse on top of the equation above -- or try maybe inserting a carriage return -- it should display nicely.)

2. Read the help page for 'plot' in the "Help" tab of the lower-right pane of your Rstudio session. See if you can re-do the graph of age-specific fertility in 1851 France, changing the color of the lines and points to "red" and changing the label on the x-axis to "Age at start of interval". (Hint: the command for how to change the labels of the axes is in the main part of the help page, but for the color you'll have to look under "Examples".)



3. Answer the following multiple choice question. If the NRR in France were found to be 0.97, does that mean that in 1851 

A. the intrinsic growth rate was less than zero

B. the observed population growth rate was less than zero

C. both of the above

```{r}
## run this section of code to reveal the answer
ans <- c("C", " ", "s", 
         "i", " ", "r", 
         "e", "w", "s",
         "n", "a", " ", 
         "e", "h", "T")
print(paste0(rev(ans), collapse = ""))
```

For more on the relationship between the NRR and population growth, section 6.3 of the Wachter book, particularly the final paragraph of this section on page 133.


