---
title: "Gravity Model Project Ideas (with Example Code)"
author: Neal Marquez
date: 30 May 2019
output: html_notebook
---

# Overview

This notebook contains instructions for the small group project as
well as a basket of ideas for possible projects. For some of these,
marked with (*), I've provided some sample code to get you started.


# Project Instructions


1) Split into groups of 4-6 people. Let's make sure we spread out
computing expertise among all of the groups. Come up with a cool team
name.

2) Come up with a project topic. 
-- Hint 1: Look at the project ideas
 listed below. 
-- Hint 2: Raise your hand and call me over to discuss
 if you want to do something else, or if you have other
 questions. 
-- Hint 3: Don't stress. All projects ideas will turn out fine --
 sometimes we learn the most from things that totally don't work
 out!

3) Write down a few sentences about what you expect to find (before
doing any analysis).

4) See if there's any sample analysis with code that would be useful
to work through in order to get coding ideas for the project your
group has chosen.

5) Get a good start on the analysis on Wednesday afternoon. We have to
leave at 4:00 pm sharp to get to the Botanical Garden.

6) On Thursday afternoon, we'll have about 2-3 more hours. In this
time, you'll want to finalize your analysis and prepare your slide
presenation.

7) The slide presentation should have about 6 slides. Here's a
possible template:
-- Slide 1: Title of project, name of group, group member names

-- Slide 2: Description of project question -- and what you expected
            to find before doing your analysis.
-- Slide 3: Your results 
           (1 slide is great, max is 3 slides of results)
-- Slide 4: What the audience should take away
-- Slide 5: What you learned (about modeling? about computing? about
            human behavior? about migration?)

# Josh's Basket of Project Ideas

-- Project idea number 1: Don't do any new modeling and just draw a
   bunch of maps and try to make sense of the migration geography of
   the United States ... What is the "view from Texas"? Or "Iowa"?

-- (*) Project idea number 2: Choose a demographic sub-population and
   see how demographic distance looks for this group. Is the "view
   from New York" (demographic distance) the same for men and women,
   for blacks and whites, for old and young, for rich and poor, for
   more and less educated, for humanities majors and business majors,
   ...?  (Hint: do some predictions before you do your
   analysis). Compare different sub-populations, but not too
   many. Explore what happens if you change Mj and Mi so that they
   reflect the sub-populations.

-- (*) An even more specific idea 3: Define Fij as moves by
   African-Americans, Mi as number of African-Americans in origin
   state, and Mj as total population in destination state. Create a
   new variable: % African-American in each state. Add this to the
   regression and see if it's a predictor. Do you think it will be? Is
   your expectation fulfilled? How do we interpret the regression
   output?

-- (*) Project idea 4: Does unemployment drive migration? Include it
   in the gravity model. We can calculate unemployment rates for each
   state and merge them with our gravity model data. Then we add
   unemployment in origin, or destination, or the gap, as as a
   covariate. See if the effect of unempoloyment depends on whether we
   use the gravity model.

-- Project idea 5: Include some other state-level covariate that you
   hypothesize might origin and destinations for migrants.

Even more ideas:

-- 6. Develop a new measure of distance -- e.g. number of border
   crossings and see if they matter. (Not sure how to calculate the
   number of border crossings using the computer!)

-- 7. See if regions matter (use census regions and have Rij matrix
   that takes value 1 if in same region and 0 if not)

-- 8. Figure out more about interpreting the gravity model. Why is
   coefficient on Mi < 1; why is coefficient on Mj < 1?  Could it have
   something to do with big states having more square miles?

-- 9. Does distance mean something different East and WEst of the
   Mississippi? Could one answer this by looking at the coefficient on
   distance in the gravity model?

# Sample Code

Here I provide some examples that I explored. The idea is not to tell
you what you must explore. Rather, the goal is to provide you with
some functioning computer code for doing your own analysis.

Not a bad idea to restart R from within RStudio (Session-->Restart R)
and load everything up from scratch
```{r}
source("project_modeling.R")
```

## (*) Project idea number 2: Views of other groups

Choose a demographic sub-population and see how demographic distance
looks for this group. Is the "view from New York" (demographic
distance) the same for men and women, for blacks and whites, for old
and young, for rich and poor, for more and less educated, for
humanities majors and business majors, ...?  (Hint: do some
predictions before you do your analysis). Compare different
sub-populations, but not too many. Explore what happens if you change
Mj and Mi so that they reflect the sub-populations vs. the total
population.

For example, we might hypothesize that California would be especially
attractive for New Yorkers who are college graduates. To study this,
let's fit the gravity model to college graduates only.

### Prepare the data
```{r}
## prepare data
table(dt.clean$educ) ## look at educational categories
## define a collgrad variabel
dt.clean[, collgrad := educ %in%
               c("4 years of college",
                 "5+ years of college")] ## creates a variable to select on
## create gravity data with select sub-populations
new.grav.dt <- create.gravity.data(my.dt = dt.clean[collgrad == TRUE,])

## NOTE: you can change the analysis by redefining this step. For
## example, if you want to only look at those over age 65
## > new.grav.dt <- create.gravity.data(my.dt = dt.clean[as.numeric(age) > 65,])

nrow(new.grav.dt) ## should be 50x50 = 2500 rows
new.grav.dt <- add.distance.variable.to.gravity.data(grav.dt = new.grav.dt)
print(new.grav.dt) ## check if distance column is there
```

### Re-estimate model

We use the new data set new.grav.dt and re-estimate the gravity model.

```{r}
new.m.pois <- glm(
    Fij ~ 1 + log(Mi) + log(Mj) + log(distance),
    family = poisson(link = "log"),
    data = new.grav.dt)
mtable(new.m.pois, summary.stats = F)
```

Generate predicted migration numbers

```{r}
## generate predictions
new.grav.dt$log.Fij.hat <- predict(new.m.pois, newdata = new.grav.dt)
new.grav.dt$Fij.hat <- exp(new.grav.dt$log.Fij.hat)
print(new.grav.dt)
## let's clean up just a bit to keep easy to read
new.grav.dt$log.Fij.hat <- NULL
new.grav.dt$Fij.hat <- round(new.grav.dt$Fij.hat, 1)
print(new.grav.dt)
```

```{r}
## re-look at demographic distance
new.grav.dt$demographic.distance <- round(new.grav.dt$Fij/new.grav.dt$Fij.hat,2)
## let's re-look at New York destinations
new.grav.dt[origin == "new york", .(destination, demographic.distance)][order(demographic.distance)]
## create vector for mapping
new.demographic.distance.ny.vec <- new.grav.dt[origin == "new york"]$demographic.distance
names(new.demographic.distance.ny.vec) <- new.grav.dt[origin == "new york"]$destination

draw.usa.map(new.demographic.distance.ny.vec,
             main = "'Demographic Distance' of Destinations from New York\n Ratios of observed migration to prediction from gravity model \n college grads only")
```

> (Josh's note: perspective for college grads looks suprisingly similar as for everyone.)

## (*) An even more specific idea 3

Define Fij as moves by African-Americans, Mi as number of
African-Americans in origin state, and Mj as total population in
destination state. Create a new variable: % African-American in each
state. Add this to the regression and see if it's a predictor. Do you
think it will be? Is your expectation fulfilled? How do we interpret
the regression output?

```{r}
## view race codes
table(dt.clean$race)
## define a True/False variable for being African American
dt.clean$black <- dt.clean$race == "black/african american/negro"
table(dt.clean$black)
new.grav.dt <- create.gravity.data(my.dt = dt.clean[black == TRUE,])  ## gravity data with selected sup-population
```

Up until now, this is like the previous example, but now we're going
to do something new. We are going to use Mj for total pop. So we
recreate grav.dt for the total population and merge with the new data
set.

```{r}
## recreate grav.dt for the total pop
grav.dt <- create.gravity.data(my.dt = dt.clean)
grav.dt <- add.distance.variable.to.gravity.data(grav.dt = grav.dt)

## now we merge the two data sets
merge.grav.dt <- merge(x = grav.dt,
                       y = new.grav.dt,
                       by = c("origin", "destination"))
## view and see that the names of the variables have suffixes .x and .y
print(merge.grav.dt)
## now define our gravity model variables the way we want
merge.grav.dt$Fij <- merge.grav.dt$Fij.y ## blacks
merge.grav.dt$Mi <- merge.grav.dt$Mi.y # blacks
merge.grav.dt$Mj <- merge.grav.dt$Mj.x #total

## create a variable for percentage black in the destination
merge.grav.dt$perc.black.dest <- 100 *merge.grav.dt$Mj.y/merge.grav.dt$Mj.x

## check by displaying
merge.grav.dt[, .(origin, destination, perc.black.dest)]
```

### Re-fit the gravity modelmodel

We now re-fit the gravity model to our newly defined Mi, Mj, Fij,
 together with the new variable perc.black.dis

```{r}
m.pois.black <- glm(
    Fij ~ 1 + log(Mi) + log(Mj) + log(distance) + perc.black.dest,
    family = poisson(link = "log"),
    data = merge.grav.dt)
mtable(m.pois, m.pois.black, summary.stats = F)
```

## Food for thought
> 1. Looks like perc black matters but how do we units of coefficients
>    in terms of elasticity?
> 2. How could a map help us understand what's going on?
> 3. What are some alternative modeling strategies? Does it make sense
>    to use the total population as the Mj? 



## (*) Project idea 4:  Does unemployment drive migration? 

Include unemployment in the gravity model. We can calculate
unemployment rates for each state and merge them with our gravity
model data. Then we add unemployment in origin, or destination, or the
gap, as as a covariate. See if the effect of unempoloyment depends on
whether we use the gravity model.

```{r}
## look at categories for employment status
table(dt.clean$empstat)

## calculate UE rate by state
ue.tmp <- dt.clean[, sum(empstat == "unemployed") /
                     (sum(empstat == "unemployed") +
                      sum(empstat == "employed")),
                   by = origin]
## assign to a single vector
ue.vec <- ue.tmp$V1
names(ue.vec) <- ue.tmp$origin

## re-assign back to our gravity data
grav.dt$origin.ue <- ue.vec[grav.dt$origin]
grav.dt$destination.ue <- ue.vec[grav.dt$destination]
```

Now we refit the model

```{r}
destination.ue.pois <- glm(
    Fij ~ 1 + log(Mi) + log(Mj) + log(distance) + destination.ue,
    family = poisson(link = "log"),
    data = grav.dt)
mtable(m.pois, destination.ue.pois, summary.stats = "N")
```
 UE at destination doesn't seem to matter. Let's try something
 fancier. The gap between unemployment in destination and
 origin. We would expect that a negative gap would lead to more
 immigration, with people migrating to states with lower
 unemployment.
 
```{r}
##  calculate gap as : unemployment in dest - unemployment in origin
grav.dt$ue.gap <- grav.dt$destination.ue - grav.dt$origin.ue
```

Now re-estimate the model
```{r}
gap.ue.pois <- glm(
    Fij ~ 1 + log(Mi) + log(Mj) + log(distance) + ue.gap,
    family = poisson(link = "log"),
    data = grav.dt)
mtable(m.pois, destination.ue.pois, gap.ue.pois, summary.stats = F)
```

OK, this seems to be what we expected now. Negative sign here means
 positive gap --> less migration; negative gap --> more migration.
 So if virginia has low UE and west virgini has high UE, gap is
 negative, and more people would move from WV --> VA.

 Let's look at the data again, just to check.
```{r}
grav.dt[origin == "west virginia" & destination == "virginia"]
##           origin destination distance  N Fij    Mi    Mj  origin.ue
## 1: west virginia    virginia 2.610803 40  40 17727 83631 0.06044609
##    destination.ue      ue.gap
## 1:     0.03980423 -0.02064186
## negative coefficient on gap means we would predict MORE migration
## from WV-->VA, which makes sense.
```

Let's redo for origin unemployment
```{r}
origin.ue.pois <- glm(
    Fij ~ 1 + log(Mi) + log(Mj) + log(distance) + origin.ue,
    family = poisson(link = "log"),
    data = grav.dt)
mtable(m.pois, destination.ue.pois, gap.ue.pois, origin.ue.pois,
       summary.stats = F)
```
And separate origin and destination effects
```{r}
dest.origin.ue.pois <- glm(
    Fij ~ 1 + log(Mi) + log(Mj) + log(distance) + origin.ue + destination.ue,
    family = poisson(link = "log"),
    data = grav.dt)
mtable(m.pois, destination.ue.pois, gap.ue.pois, origin.ue.pois, dest.origin.ue.pois,
       summary.stats = F)
```

It looks like, in this analysis, that it's only the unemployment rate
in the origin state that matters. Why might this be?

### Last example with unemployment. 

Let's see how much it matters that we're using gravity model with
distance and size of destination. We can do this be reestimating
effect of origin.ue on a model that leaves out distance and
destination pop size.

```{r}

origin.ue.reduced<- glm(
    Fij ~ 1 + log(Mi) + origin.ue, ## note: we have no log(Mj) and no log(distance)
    family = poisson(link = "log"),
    data = grav.dt)
mtable(origin.ue.pois, origin.ue.reduced, summary.stats = "N")
```

Wow. This seems to give us some motivation for using the gravity
model. If we didn't include distance and size of destination, and only
looked at out-migration rates by unemployment level, we would think
that unemployment had little or no effect.

# Congratulations!!! 

> You've scrolled (and perhaps even worked part of the way) to the end
> of example code for the small group projects.



